Hydrous melting and partitioning in the deep upper mantle: insights from experiments in the system MgO-SiO$_2$-H$_2$O

### 2015-07-08 MSH.tex
Update

Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that the thermodynamic properties of the melt remain essentially constant between 13 and 20 GPa. Water concentrations in the solid are taken from the original studies. We restrict our analysis to data from 1100--1400 C, where the Mg:Si ratios of the melts are similar to that in forsterite. The average $D^{wad/melt}_{H_2O}$ (0.44) and $D^{rw/melt}_{H_2O}$ values (0.6) produce the lines in Figure \ref{fig:partitioning}. Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone; the opposite conclusion to that reached by \cite{DDFK2005} using the same data XXXX check this!!. This prediction could explain observations of low velocity layers at ca. 520 km depth \citep{JDH2010}, in an extension of the water filter model of \cite{BK2003}.

An accurate thermodynamic model for hydrous melts can also reveal the mechanism of hydrogen incorporation into nominally anhydrous minerals. This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid: XXXX correct this!!
\begin{equation}
  K = \frac{a_{OH}^n}{a_{H_2O}^{n/2}a_O^{n/2}}
\end{equation}

Isolated protons (n=0.5), proton pairs (n=1) and proton . Making the approximation that $a_O = 1$, which is probably reasonable at low H concentrations in the solid 
Making the assumption that concentration of H$_2$O in the solid is proportional to the activity of OH in the melt, we therefore have
\begin{equation}
  c \propto a_{H_2O}^n
\end{equation}
where $n$ relates to the mechanisms of hydrogen incorporation into the solid \citep{KB2006}

>>>>

Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that the thermodynamic properties of the melt remain essentially constant between 13 and 20 GPa. Water concentrations in the solid are taken from the original studies. We restrict our analysis to data from 1100--1400 C, where the Mg:Si ratios of the melts are similar to that in forsterite. The average $D^{wad/melt}_{H_2O}$ (0.44) and $D^{rw/melt}_{H_2O}$ values (0.6) produce the lines in Figure \ref{fig:partitioning}. Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone. This prediction could explain observations of low velocity layers at ca. 520 km depth \citep{JDH2010}, in an extension of the water filter model of \cite{BK2003}. We also note that there is currently no strong evidence to argue for repartitioning of water between the upper and lower transition zone due to the secular cooling of the Earth as proposed by \cite{DDFK2005}.

An accurate thermodynamic model for hydrous melts can also reveal the mechanism of hydrogen incorporation into nominally anhydrous minerals. This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid: 
\begin{equation}
  K = \frac{a_{(OH)_{2n}}}{a_{H_2O}^{n}a_O^{n}}
\end{equation}
The value of n depends on the H$^+$ substitution mechanism: isolated protonation yields n=0.5, proton pairs n=1 and 4-proton clusters n=2. Making the approximation that $a_O$ is constant, that the concentration of OH in the solid is proportional to the activity of (OH)$_{2n}$, and that the equilibrium constant has the usual Arrhenius form \citep{KB2006}, we therefore have
\begin{equation}
  c = A a_{H_2O}^{n} \exp \left( -\frac{\Delta H^{1 bar} + \Delta V^{solids} P}{R T}\right)
\end{equation}
